---
title: "Project03"
author: "Kaylie Evans"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary:
1- find what rows are missing salary ranges 
2- find what fields have salary information
3- split the salary column into 2 columns for range
4- move salary information for the rows that contain the information in other rows
5- reformat the hourly wages to salary numbers


## Load Libraries:
```{r library-load}
library(tidyverse)
library(dplyr)
library(stringr)
```


#To be adjusted:
## Load .csv:
```{r csv-load}
#reading in the data set to an object named after the inflation data set (infl)
jobs_raw <- read.csv("https://raw.githubusercontent.com/naomibuell/DATA607_Project3/main/linkedin-jobs-usa.csv", sep = ',')

#preserve the raw file
jobs <- jobs_raw

#preview data
glimpse(jobs)
```


## 1
### Find what rows are missing salary information
```{r 1}
#only missing salaries
salary_finder_02 <- jobs |> 
  filter(
    salary == NaN
  ) 
```


## 2
### Find what fields have salary ranges 
```{r 2}
#title
count(
  salary_finder_02 |>
  mutate(
    title = str_extract_all(title, "salary") 
    ) |>
        filter(
          title == "salary"
        )
)

#company
count(
  salary_finder_02 |>
  mutate(
    company = str_extract_all(company, "salary") 
  ) |>
      filter(
        company == "salary"
      )
)

#description
count(
  salary_finder_02 |>
  mutate(
    description = str_extract_all(description, "salary") 
  ) |>
      filter(
        description == "salary"
      )
)

#location
count(
  salary_finder_02 |>
  mutate(
    location = str_extract_all(location, "salary") 
  ) |>
      filter(
        location == "salary"
      )
)

#criteria
count(
  salary_finder_02 |>
  mutate(
    criteria = str_extract_all(criteria, "salary") 
  ) |>
      filter(
        criteria == "salary"
      )
)
```

## 3
### split the salary column into 2 columns for range
```{r 3}
jobs <- jobs |>
  mutate(
    sal_high = str_extract(salary, "(?<=-).*"),
    .after = salary
  ) |>
      mutate(
        sal_low = str_extract(salary, ".*(?=-)"),
        .before = sal_high
      ) 
```


## 4
### move salary information for the rows that contain the information in other rows
```{r 4}
jobs <- jobs |> 
  mutate(
    key_field = c(1:2845),
    .before = title
  ) 

#rows where salary information is already available
salary_exist <- jobs |> 
  filter(
    salary != NaN
  ) 

#rows where there is no salary information
salary_finder_01 <- jobs |> 
  filter(
    salary == NaN
  ) 

#filling all available salaries from description
salary_finder_03 <- salary_finder_01 |>
  mutate(
    description_02 = str_extract_all(description, "(Salary.*\\d+[,]\\d+)|(salary.*\\d+[,]\\d+)"),
    .before = "description"
  ) |>
      filter(
        description_02 != "character(0)"
      ) |> 
          mutate(
            description_03 = as.character(str_extract_all(description_02, "\\d+[,]\\d+")),
            .before = "description"
          ) |>
              mutate(
                sal_low = str_extract(description_03, "\\d+[,]\\d+"),
                sal_high = str_extract(description_03, "(?<=(\\s)).*"),
                sal_high = str_extract(sal_high, "\\d+[,]\\d+"),
                sal_high = ifelse(is.na(sal_high), sal_low, sal_high),
                salary = paste(sal_low, " - ", sal_high),
                description_02 = NULL,
                description_03 = NULL
              )

#rows where there was no salary information available
salary_finder_02 <- salary_finder_01 |>
  mutate(
    description_02 = str_extract_all(description, "(Salary.*\\d+[,]\\d+)|(salary.*\\d+[,]\\d+)"),
    .before = "description"
  ) |>
      filter(
        description_02 == "character(0)"
      ) |>
          mutate(
            description_02 = NULL
          )

#joining all rows back together
jobs_salary <- rbind(salary_exist, salary_finder_02, salary_finder_03)
jobs_salary <- jobs_salary[order(jobs_salary$key_field), ]
```


## 5
### reformat the hourly wages to salary numbers
```{r 5}
jobs_clean <- jobs |>
  mutate(
    sal_low = as.numeric(str_extract_all(sal_low, "//d+")),
  ) 

sal_monthly <- jobs_clean |>
      filter(
        sal_low > 100,
        sal_low < 3000
      ) |>
          mutate(
            sal_low = sal_low * 12
          )

#less than 3000 and greater than 100: x12
#less than 300: x2080
```
